Lab 11 - Missing Data and Imputation:Answers

### 2023-04-05 Lab 11 Missing data/Lab11_MissingData.Rmd
Add files via upload

**Data!**
First, let's get our data set up. 
```{r}
#clear global environ
rm(list=ls())

#load in airquality data
data_clean <- airquality
summary(data_clean)

#here's our "true model" for future reference
model_true <- lm(Temp~ Ozone + Solar.R + Wind, data = data_clean)

#lets copy this df and then add some missingness
data_miss <- data_clean
data_miss[4:9,3] <- rep(NA,6)
data_miss[1:8,4] <- NA
summary(data_miss) #check out those *~* NAs ~*~

@@

Alright, let's take a look at those NA's more deeply
```{r}
#what does this do?
is.na(data_miss$Ozone)

#and this?
which(is.na(data_miss$Ozone))

#and this?
sum(is.na(data_miss$Ozone))

@@

#let's create an lm with this model to compare to later iterations
#model Temp as your y, Ozone, Solar.R, and Wind as your x values
model_listwise <- lm(Temp ~ Ozone + Solar.R + Wind, data = data_listwise)

@@

#uses if else statements to recode NA values as the mean of the rest of that column
#is.an()
data_recode$Ozone <- ifelse(is.na(data_recode$Ozone), 
                            mean(data_recode$Ozone, na.rm = TRUE),
                            data_recode$Ozone)

data_recode$Solar.R<- ifelse(is.na(data_recode$Solar.R),
                            mean(data_recode$Solar.R, na.rm = TRUE),
                            data_recode$Solar.R)

data_recode$Wind<- ifelse(is.na(data_recode$Wind),
                            mean(data_recode$Wind, na.rm = TRUE),
                            data_recode$Wind)

data_recode$Temp <- ifelse(is.na(data_recode$Temp),
                            mean(data_recode$Temp, na.rm = TRUE),
                            data_recode$Temp)
#summarize
summary(data_recode)

#linear model (same as above, but rename "model_recode")
model_recode <- lm(Temp~ Ozone + Solar.R + Wind, data = data_recode)

@@

So this is great, but then what do we do if we want to run a model? We have five data sets. One option is to pool the data this way:
```{r, eval = FALSE}
#pooling
model_mice <- with(mice_data,lm(Temp~ Ozone + Solar.R + Wind))
summary(pool(model_mice))

@@

Great, let's try it out. Examine the function by calling `?amelia()` - look at the different arguments. Notice that amelia can handle time series data uniquely. Let's factor that in by creating a running time variable.

@@

#create running time variable, lets just assume we're in 2023
data_miss$time <- make_datetime(month = data_miss$Month,
                                day = data_miss$Day,
                                year = 2023)

#impute
amelia_data <- amelia(x = data_miss,
                 m = 5,
                 time = "time")

@@

for(i in 1:5){
  models_amelia[[i]] <- lm(Temp ~ Ozone + Solar.R + Wind,
                        data=amelia_data$imputations[[i]]) #why do we index like this?
}

>>>>

**Data!**
First, let's get our data set up. 
```{r}
#clear global environment
rm(list=ls())

#load in iris data
data_clean <- iris
summary(data_clean)

#Checking number of NAs
sum(is.na(data_clean))

#here's our "true model" for future reference
model_true <- lm(Sepal.Length~ Sepal.Width+Petal.Length+Petal.Width, data = data_clean)

#lets copy this df and then add some missingness
data_miss <- data_clean
data_miss[6:27,2] <- rep(NA,2)
data_miss[102:119,4] <- rep(NA,2)
data_miss[35:50,3] <- rep(NA,2)
summary(data_miss) #check out those *~* NAs ~*~

sum(is.na(data_miss))

@@

Alright, let's take a look at those NA's more deeply
```{r}
#what does this do?
is.na(data_miss$Sepal.Width)

#and this?
which(is.na(data_miss$Sepal.Width))

#and this?
sum(is.na(data_miss$Sepal.Width))

@@

#let's create an lm with this model to compare to later iterations

model_listwise <- lm(Sepal.Length~ Sepal.Width+Petal.Length+Petal.Width, data = data_listwise)

@@

#uses if else statements to recode NA values as the mean of the rest of that column
data_recode$Sepal.Width <- ifelse(is.na(data_recode$Sepal.Width),
                            mean(data_recode$Sepal.Width, na.rm = TRUE),
                            data_recode$Sepal.Width)

data_recode$Petal.Length<- ifelse(is.na(data_recode$Petal.Length),
                            mean(data_recode$Petal.Length, na.rm = TRUE),
                            data_recode$Petal.Length)

data_recode$Petal.Width<- ifelse(is.na(data_recode$Petal.Width),
                            mean(data_recode$Petal.Width, na.rm = TRUE),
                            data_recode$Petal.Width)

data_recode$Sepal.Length <- ifelse(is.na(data_recode$Sepal.Length),
                            mean(data_recode$Sepal.Length, na.rm = TRUE),
                            data_recode$Sepal.Length)

#summarize
summary(data_recode)

#linear model (same as above, but rename "model_recode")
model_recode <- lm(Sepal.Length~ Sepal.Width+Petal.Length+Petal.Width, data = data_recode)

@@

So this is great, but then what do we do if we want to run a model? We have five data sets. One option is to pool the data this way:
```{r, eval = FALSE}
#pooling
model_mice <- with(mice_data,lm(Sepal.Length~ Sepal.Width+Petal.Length+Petal.Width))
summary(pool(model_mice))

@@

Great, let's try it out. Examine the function by calling `?amelia()` - look at the different arguments.

@@

#impute
amelia_data <- amelia(x = data_miss,
                      noms="Species",
                       m = 5)

@@

for(i in 1:5){
  models_amelia[[i]] <- lm(Sepal.Length~ Sepal.Width+Petal.Length+Petal.Width,
                        data=amelia_data$imputations[[i]]) #why do we index like this?
}

### 2023-04-05 Lab 11 Missing data/Lab11_MissingData_AnswerKey.Rmd
Add files via upload

#load in airquality data
data_clean <- airquality
summary(data_clean)

#here's our "true model" for future reference
model_true <- lm(Temp~ Ozone + Solar.R + Wind, data = data_clean)

#lets copy this df and then add some missingness
data_miss <- data_clean
data_miss[4:9,3] <- rep(NA,6)
data_miss[1:8,4] <- NA
summary(data_miss) #check out those *~* NAs ~*~

@@

#let's create an lm with this model to compare to later iterations
#model Temp as your y, Ozone, Solar.R, and Wind as your x values
model_listwise <- lm(Temp~ Ozone + Solar.R + Wind, data = data_listwise)

@@

#uses if else statements to recode NA values as the mean of the rest of that column
data_recode$Ozone <- ifelse(is.na(data_recode$Ozone),
                            mean(data_recode$Ozone, na.rm = TRUE),
                            data_recode$Ozone)

data_recode$Solar.R<- ifelse(is.na(data_recode$Solar.R),
                            mean(data_recode$Solar.R, na.rm = TRUE),
                            data_recode$Solar.R)

data_recode$Wind<- ifelse(is.na(data_recode$Wind),
                            mean(data_recode$Wind, na.rm = TRUE),
                            data_recode$Wind)

data_recode$Temp <- ifelse(is.na(data_recode$Temp),
                            mean(data_recode$Temp, na.rm = TRUE),
                            data_recode$Temp)

@@

#linear model (same as above, but rename "model_recode")
model_recode <- lm(Temp~ Ozone + Solar.R + Wind, data = data_recode)

@@

So this is great, but then what do we do if we want to run a model? We have five data sets. One option is to pool the data this way:
```{r, eval = TRUE, results='asis'}
#pooling
model_mice <- with(mice_data,lm(Temp~ Ozone + Solar.R + Wind))

@@

#create running time variable, lets just assume we're in 2023
data_miss$time <- make_datetime(month = data_miss$Month,
                                day = data_miss$Day,
                                year = 2023)

#impute
amelia_data <- amelia(x = data_miss,
                 m = 5,
                 time = "time")

@@

#alternatively!
all_coefs <- sapply(models_africa, coef) #get coefs
all_ses <- sapply(models_africa, function(x) sqrt(diag(vcov(x)))) #get ses

mi.meld(q = all_coefs, se = all_ses, byrow = F) 
```

>>>>

#load in iris data
data_clean <- iris
summary(data_clean)

#Checking number of NAs
sum(is.na(data_clean))

#here's our "true model" for future reference
model_true <- lm(Sepal.Length~ Sepal.Width+Petal.Length+Petal.Width, data = data_clean)

#lets copy this df and then add some missingness
data_miss <- data_clean
data_miss[6:27,2] <- rep(NA,2)
data_miss[102:119,4] <- rep(NA,2)
data_miss[35:50,3] <- rep(NA,2)
summary(data_miss) #check out those *~* NAs ~*~

sum(is.na(data_miss))

@@

#let's create an lm with this model to compare to later iterations

model_listwise <- lm(Sepal.Length~ Sepal.Width+Petal.Length+Petal.Width, data = data_listwise)

@@

#uses if else statements to recode NA values as the mean of the rest of that column
data_recode$Sepal.Width <- ifelse(is.na(data_recode$Sepal.Width),
                            mean(data_recode$Sepal.Width, na.rm = TRUE),
                            data_recode$Sepal.Width)

data_recode$Petal.Length<- ifelse(is.na(data_recode$Petal.Length),
                            mean(data_recode$Petal.Length, na.rm = TRUE),
                            data_recode$Petal.Length)

data_recode$Petal.Width<- ifelse(is.na(data_recode$Petal.Width),
                            mean(data_recode$Petal.Width, na.rm = TRUE),
                            data_recode$Petal.Width)

data_recode$Sepal.Length <- ifelse(is.na(data_recode$Sepal.Length),
                            mean(data_recode$Sepal.Length, na.rm = TRUE),
                            data_recode$Sepal.Length)

@@

#linear model (same as above, but rename "model_recode")
model_recode <- lm(Sepal.Length~ Sepal.Width+Petal.Length+Petal.Width, data = data_recode)

@@

So this is great, but then what do we do if we want to run a model? We have five data sets. One option is to pool the data this way:
```{r, eval = TRUE, results='asis'}
#pooling
model_mice <- with(mice_data,lm(Sepal.Length~ Sepal.Width+Petal.Length+Petal.Width))

@@

#imputing (setting factor to nominal variable)
amelia_data <- amelia(x = data_miss,
                      noms="Species",
                       m = 5)

@@
